References for PloS paper

### 2014-01-14 image_analysis_paper/plos_draft.tex
Updated draft

We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
%TODO: check that NC14 is correct
%
More specifically, we are interested in the activation dynamics of dpERK during NC14.
%
We collect fluorescent images of {\em Drosophila} embryos that have been stained for dpERK. 
%
Because we must fix the embryos before staining for dpERK, we cannot obtain a continuous time trajectory of dpERK in one embryo; instead, we obtain a set of snapshots, one per embryo, of embryos stained for dpERK. 
% 
We would like to order the resulting images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
%
% TODO: add motivation about why we are interested in this
%
Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
%
We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
%
We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.

\begin{figure}[!ht]
\begin{center}
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.65\textwidth]{drosophila_imaging_setup}
\caption{}
\label{subfig:imaging_device}
\end{subfigure}
\begin{subfigure}{0.55\textwidth}
\includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
\includegraphics[width=0.25\textwidth]{drosophila_membrane}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_dpERK}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_DI}
\caption{}
\end{subfigure}
\begin{subfigure}{0.35\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{circle_profile}\\
\includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
\caption{}
\end{subfigure}\\
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_ordered_membrane}
\caption{}
\end{subfigure}
\end{center}
\caption{{\bf Data collection.} (a) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
(b) Concentration profile of dpERK extracted from a fluorescent image. The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
(c) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(d) Concentration profiles of dpERK for many embryos from (c), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
\label{fig:background}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{coeff_12_new}
\caption{}
\label{subfig:PCA_12}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_DMAPS}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{DMAPS_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve. 
(b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
(c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
\label{fig:DMAPS_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_aligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_angsynch}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned using angular synchronization.
(c) Concentration profiles of dpERK aligned using angular synchronization and ordered by the first (non-trivial) DMAPS embedding coordinate.}
\label{fig:angsynch_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_vdm}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{VDM_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles using vector diffusion maps.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned and ordered using vector diffusion maps.
(c) Correlation between the VDM embedding coordinate and the membrane thickness.}
\label{fig:vdm_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{vdm_2d_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{dpERK_vdm_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the VDM embedding coordinate and the membrane thickness. 
(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. The corresponding data points are circled in (a). }
\label{fig:vdm_image_ordering}
\end{figure}

@@

Because these images are much more complex than the dpERK fluorescent images, we expect any alignment results to be poor.
%
Using VDM to align and order the images yielded a Spearman correlation coefficient of 0.3525.
%
However, the scattering transform is robust to small deformations, and so we expect using DMAPS with the scattering transform coefficients to give better results.
%
We computed the scattering transform coeffficients of the edge-detected images, and used DMAPS to order the images in time. 
%
The results are shown in Figure~\ref{fig:scattrans_membrane_ordering}.
%
The correlation between $\phi_2$ and the membrane thickness is not as strong as the correlations for the dpERK images; the Spearman correlation is 0.8403.
%
However, these results are still promising considering the relative simplicity of the methods. 
%
%$\rho = 0.8403$, 300 seconds for scat + DMAPS
%$\rho = 0.3525$ for VDM + membrane pics, 1100 seconds

\begin{figure}[!ht]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{DMAPS_membrane_scat_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{membrane_scat_\n}
\hfill}
\newline
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{membrane_scat_raw_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of membrane proteins.}
(a) Correlation between the seventh (non-trivial) DMAPS embedding coordinate and the membrane thickness. DMAPS was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images. The first six non-trivial DMAPS coordinates were found to capture other features of the images, such as  differences in fluorescence intensity within the images.
(b) Some representative edge-detected images (top) and fluorescent images (bottom), ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. The corresponding data points are circled in (a).}
\label{fig:scattrans_membrane_ordering}
\end{figure}

@@

Let $X \in \mathbb{R}^{n \times d}$ denote our data set, consisting of $n$ points in $d$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
%
We first mean-center the data set to produce the matrix $\hat{X}$, where
\begin{equation}
\hat{X}_{ij} = X_{ij} - \frac{1}{n} \sum_{k=1}^n X_{kj}.
\end{equation}
%
The empirical covariance of the data set, $C \in \mathbb{R}^{d \times d}$ is then given by
\begin{equation}
C = X^T X.
\end{equation}
%
We compute the eigenvectors $\psi_1, \dots, \psi_d$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d$.
%
The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_d$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_d$ measure the variability captured by the corresponding eigenvector. 
%
If there is a {\em spectral gap} such that the first few eigenvectors capture a significant portion of the variance in our data set, projection onto these first few principal components will retain most features of the data set while reducing the dimensionality.
%
In our setup, we assume that the data is inherently one-dimensional, and that projection onto $\psi_1$ will uncover the correct time ordering of the data.

@@

Given data points $x_1, \dots, x_n \in \mathbb{R}^d$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
\begin{equation}
W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon} \right)
\end{equation}
and $\epsilon$ is a characteristic distance between data points.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
%
We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$. 
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
%
The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; the next few eigenvectors give the ``meaningful'' embedding coordinates for the data.

@@

When the data set is invariant under an underlying symmetry group (such as rotations), we must first factor out the underlying symmetries before doing further analysis.
%
Angular synchronization calculates the optimal alignments for a data set, assuming we have estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.

Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
%
We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
%
Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
\begin{equation}
R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
\end{equation}
%
We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
%TODO: Check that this order is correct
%
We would also like to exploit {\em higher-order} consistency relations;
for example, the relationship $R_{ij} R_{jk} R_{ki} = I$, where $I$ is the identity matrix, should hold.
%
Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.

We define the objective function 
\begin{equation} \label{eqn:angsynch_obj}
\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j.
\end{equation}
%
%TODO: check if a norm is needed
This objective function is large if $R_i$ adn $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
%
Therefore, we are seeking the most consistent set of global alignments, given the computed pairwise alignments.

In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
%
Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
%
We then define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$).
%
We then compute the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
%
The solution to the relaxed problem is given by $\hat{R}$, so that the $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
%
This formulation also accounts for higher-order relations.
%
For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j$. 
%
However, the solution to this problem is also given by the top eigenvector of $H$.

@@

Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise.
%
However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.

Vector diffusion maps (VDM) couples the symmetry-factoring of angular synchronization with the uncovering of dynamics of diffusion maps \cite{singer2012vector}. 
%
We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
%
We then construct the block matrix $S$, where the blocks of $S$ are defined by
\begin{equation}
S_{ij} = \left\{ \begin{array}{l l} 
A_{ij} H_{ij} & i \ne j \\
0_{d \times d} & i = j
\end{array}
\right.
\end{equation}
%
where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.

We then compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
%
As in angular synchronization, the first ``block'' eigenvector, i.e. the first two eigenvectors concatenated into a $nd \times d$ matrix, gives the optimal rotations. 
%
However, the eigenvectors now also give us embedding coordinates.
%
Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$. 
%
The embedding coordinates are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.

@@

The scattering transform consists of a wavelet transform of the signal, followed by  the appropriate averaging operator.
%
This transform is applied to the signal itself, as well as to the residual, in order to preserve as much information about the signal itself, while still keeping the transform coefficients translation- and rotation-invariant. 
% 
Furthermore, the scattering transform can be shown to be a contractive operator; therefore, noise and small deformations in the signal are not amplified in the transformation.
%
We utilized the freely available ScatNet software (\url{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.

>>>>

We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
%TODO: check that NC14 is correct
%
More specifically, we are interested in the activation dynamics of the ERK pathway during NC14, which we monitor via the production of doubly phosphorylated ERK (dpERK).
%
We collect fluorescent images of {\em Drosophila} embryos that have been stained for dpERK. 
%
Because we must fix the embryos before staining for dpERK, we cannot obtain a continuous time trajectory of dpERK in one embryo; instead, we obtain a set of snapshots, one per embryo, of embryos stained for dpERK. 
% 
We would like to order the resulting images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
%
% TODO: add motivation about why we are interested in this
%
Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
%
We then use image analysis software to extract the dpERK concentration profile around the circumference of the embryo from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
%
We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.

\begin{figure}[!ht]
\begin{center}
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.65\textwidth]{drosophila_imaging_setup}
\caption{}
\label{subfig:imaging_device}
\end{subfigure}
\begin{subfigure}{0.55\textwidth}
\includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
\includegraphics[width=0.25\textwidth]{drosophila_membrane}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_dpERK}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_DI}
\caption{}
\end{subfigure}
\begin{subfigure}{0.35\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{circle_profile}\\
\includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
\caption{}
\end{subfigure}\\
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_ordered_membrane}
\caption{}
\end{subfigure}
\end{center}
\caption{{\bf Data collection.} (a) Schematic of microfluidic device used for imaging. Embryos are flown through the channels of the device, and one embryo is caught in each ``trap''. The device can then be placed in the confocal microscope for imaging. (b) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
(c) Concentration profile of dpERK extracted from a fluorescent image. The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
(d) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(e) Concentration profiles of dpERK for many embryos from (d), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
\label{fig:background}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{coeff_12_new}
\caption{}
\label{subfig:PCA_12}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_DMAPS}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{DMAPS_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve. 
(b) Concentration profiles of dpERK from \ref{subfig:unordered_profiles}, now ordered by the first non-trivial DMAPS embedding coordinate.
(c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
\label{fig:DMAPS_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_aligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_angsynch}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circular profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right (this corresponds to rotating the circular profile).
(b) Concentration profiles of dpERK aligned using angular synchronization.
(c) Concentration profiles of dpERK aligned using angular synchronization and ordered by the first (non-trivial) DMAPS embedding coordinate.}
\label{fig:angsynch_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_vdm}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{VDM_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles using vector diffusion maps.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circular profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned and ordered using vector diffusion maps.
(c) Correlation between the VDM embedding coordinate and the membrane thickness.}
\label{fig:vdm_ordering}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{vdm_2d_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{dpERK_vdm_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the VDM embedding coordinate calculated from two-dimensional fluorescent images and the membrane thickness. 
(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. The corresponding data points are circled in (a). }
\label{fig:vdm_image_ordering}
\end{figure}

@@

Because these images are much more complex than the dpERK fluorescent images, we expect any alignment results to be poor.
%
Using VDM to align and order the images yielded a Spearman correlation coefficient of 0.3525.
%
However, the scattering transform is robust to small deformations, and so we expect using DMAPS with the scattering transform coefficients to give better results.
%
We computed the scattering transform coefficients of the edge-detected images, and used DMAPS to order the images in time. 
%
The results are shown in Figure~\ref{fig:scattrans_membrane_ordering}.
%
The correlation between $\phi_2$ and the membrane thickness is not as strong as the correlations for the dpERK images; the Spearman correlation is 0.8403.
%
However, these results are still promising considering the relative simplicity of the methods. 
%
%$\rho = 0.8403$, 300 seconds for scat + DMAPS
%$\rho = 0.3525$ for VDM + membrane pics, 1100 seconds

\begin{figure}[!ht]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{DMAPS_membrane_scat_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{membrane_scat_\n}
\hfill}
\newline
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{membrane_scat_raw_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of membrane proteins.}
(a) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness. DMAPS was computed using the translation- and rotation-invariant scattering transform coefficients of the edge-detected images as ``features'' of the images.
(b) Some representative edge-detected images (top) and fluorescent images (bottom), ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. The corresponding data points are circled in (a).}
\label{fig:scattrans_membrane_ordering}
\end{figure}

@@

Let $X \in \mathbb{R}^{n \times p}$ denote our data set, consisting of $n$ points in $p$ dimensions, where $X_{ij}$ denotes the $j^{th}$ dimension of the $i^{th}$ observation. 
%
We first mean-center the data set to produce the matrix $\hat{X}$, where
\begin{equation}
\hat{X}_{ij} = X_{ij} - \frac{1}{n} \sum_{k=1}^n X_{kj}.
\end{equation}
%
The empirical covariance of the data set, $C \in \mathbb{R}^{p \times p}$ is then given by
\begin{equation}
C = \frac{1}{n-1} X^T X.
\end{equation}
%
We compute the eigenvectors $\psi_1, \dots, \psi_p$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_p$ of $C$, and order them such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_p$.
%
The principal components of the data set $X$ are given by $\psi_1, \dots, \psi_p$, and the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_p$ measure the variability captured by projecting the data onto the corresponding eigenvector. 
%
If there is a {\em spectral gap} such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_d \gg \lambda_{d+1} \ge \dots \ge \lambda_p$, then projecting the data onto the first $d$ eigenvectors will retain most of the variance of the data set, while reducing the dimensionality from $p$ to $d$.
%
In our setup, we assume that the data is inherently one-dimensional ($d=1$), and that projection onto $\psi_1$ will uncover the correct time ordering of the data.

@@

Given data points $x_1, \dots, x_n \in \mathbb{R}^p$, we fist calculate the matrix $W \in \mathbb{R}^{n \times n}$, where 
\begin{equation}
W_{ij} = \exp \left( -\frac{d^2(x_i, x)j)}{\epsilon^2} \right)
\end{equation}
and $\epsilon$ is a characteristic distance between data points.
%
$\epsilon$ can be chosen using several techniques (see, for example \cite{coifman2008graph}); in practice, we choose $\epsilon$ to be the median of the pairwise distances between data points.
%
We then compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{n} W_{ij}$, and the matrix $A = D^{-1} W$. 
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ and eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_n$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$. 
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors. 
%
The eigenvectors $\phi_1, \phi_2, \dots, \phi_n$ give the embedding coordinates, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point. 
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector.
%
The next few eigenvectors give the ``meaningful'' embedding coordinates for the data.

@@

When a data set is invariant under an underlying symmetry group (such as rotations), we must factor out the underlying symmetries before doing further analysis.
%
Angular synchronization calculates the optimal alignments for each point in a data set, given estimates of the optimal alignments between pairs of data points \cite{singer2011angular}.

Let $x_1, \dots, x_n$ denote the data points that we wish to align under some symmetry group.
%
We will consider the case where the symmetry group is the group of $d$-dimensional rotations, $SO(d)$. 
%
Let $R_{ij} \in SO(d)$ denote the rotation that aligns $x_j$ to $x_i$, so that
\begin{equation}
R_{ij} = \argmin_{R \in SO(d)} \|Rx_j - x_i \|^2.
\end{equation}
%
We seek to find the rotations $R_1, R_2, \dots, R_n \in SO(d)$ such that $R_i R_j^T \approx R_{ij}$, for every pair $i, j$. 
%TODO: Check that this order is correct
%
We would also like to exploit {\em higher-order} consistency relations;
for example, the relationship $R_{ij} R_{jk} \approx R_{ik}$ should hold if the rotation estimates are accurate.
%
Therefore, we consider measurements which (almost) satisfy such conditions ``good'' measurements, and those measurements which do not satisfy such conditions as most likely inaccurate.

We would like to solve the optimization problem 
\begin{equation} \label{eqn:angsynch_obj}
\max_{R_1, \dots, R_n \in SO(d)} \left\| \sum_{i=1}^{n} \sum_{j=1}^{n} R_i^T R_{ij} R_j \right\|^2.
\end{equation}
%
%TODO: check if a norm is needed
Each term in the objective function will be large large if $R_i$ and $R_j$ are consistent with $R_{ij}$, and will be random with mean 0 if $R_i$ and $R_j$ are inconsistent with $R_{ij}$.
%
Therefore, solving \eqref{eqn:angsynch_obj} gives us the set of global alignments which are most consistent with the computed pairwise alignments.

In general, the solution to \eqref{eqn:angsynch_obj} is not easily computed.
%
Instead, we relax the problem and allow $R_1, \dots, R_n \in \mathbb{R}^{d \times d}$.
%
We define the matrix $H$, which is an $n \times n$ block matrix with $d \times d$ blocks (so $H \in \mathbb{R}^{nd \times nd}$), with $H_{ij} = R_{ij}$.
%
We can then write the relaxed problem as 
\begin{equation} \label{eqn:angsynch_relax}
\max_{R\in \mathbb{nd \times d}} \| R^T H R \|^2.
\end{equation}
%
The solution to \eqref{eqn:angsynch_relax} is given by the top ``block'' eigenvector of $H$, which we denote $\hat{R} \in \mathbb{R}^{nd \times d}$. 
%
The $i^{th}$ rotation is (approximately) $\hat{R}(i) \in \mathbb{R}^{d \times d}$.
%
We would like to note that this formulation also accounts for higher-order relations.
%
For example, if we want to optimize over all pairs of pairwise rotations, we would solve $\max_{R_1, \dots, R_n \in SO(d)} \sum_{i=1}^{n} \sum_{j=1}^{n} \left\| R_i^T \left( \sum_k R_{ik} R_{kj} \right) R_j \right\| = \| R^T H^2 R \|^2$. 
%
However, the solution to this problem is also given by the top eigenvector of $H$.

@@

Angular synchronization assumes that each data point is a replicate measurements of the same underlying configuration, simply corrupted with rotations and some noise, and therefore considers all pairwise measurements equally important.
%
However, in system with an underlying symmetry group {\em as well as} a dynamical process, the data points are not identical modulo symmetries and noise, and we would like to place more emphasis on aligning data points which are dynamically close.

Vector diffusion maps (VDM) couples the symmetry-reduction of angular synchronization with the automatic parameterization of diffusion maps \cite{singer2012vector}. 
%
We construct the matrix $A$ as in diffusion maps, and the matrix $H$ as in angular synchronization.
%
We then construct the block matrix $S$, where the blocks of $S$ are defined by
\begin{equation}
S_{ij} = \left\{ \begin{array}{l l} 
A_{ij} H_{ij} & i \ne j \\
0_{d \times d} & i = j
\end{array}
\right.
\end{equation}
%
where $A_{ij}$ is the $(i,j)$ entry of $A$, and $H_{ij}$ is the $(i,j)$ block of $H$.

We compute the eigenvectors $\phi_1, \dots, \phi_{nd}$ and eigenvalues $\lambda_1, \dots, \lambda_{nd}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{nd}|$.
%
As in angular synchronization, the first ``block'' eigenvector, i.e. the first $d$ eigenvectors concatenated into a $nd \times d$ matrix, gives the (relaxed) optimal rotations. 
%
However, the eigenvectors now also give us embedding coordinates.
%
Let $\phi_j(i)$ denote the $i^{th}$ block of $\phi_j$ (so that $\phi_j(i) \in \mathbb{R}^d$). 
%
The embedding coordinates of $x_i$ are then given by $\langle \phi_j(i), \phi_k(i) \rangle$, where $1 \le j, k, \le nd$.

@@

The scattering transform consists of a wavelet transform of the signal, followed by  the appropriate averaging operator.
%
This transform is applied to the signal itself, as well as to the residual, in order to preserve as much information about the signal itself, while still keeping the transform coefficients translation- and rotation-invariant. 
% 
Furthermore, the scattering transform can be shown to be a contractive operator, so that noise and small deformations in the signal are not amplified in the transformation.
%
We utilized the freely available software ScatNet (\url{http://www.di.ens.fr/data/software/scatnet/}) to compute the rotation-invariant scattering coefficients.
